---
title: "ProblemSet2_new"
author: "Ke-li Chiu & Diego Mamanche Castellanos"
date: "06/02/2020"
output: pdf_document
abstract: "Education is a contributor to many beneficial socio-economic outcomes. We intend to exam the the relationship between academic majors and household income in Toronto's neighbourhoods—can we find higher or lower percentage population in certain academic majors in higher or lower income neighbourhoods? The result shows that higher income neighbourhoods also have higher percentage of residents with majors in business, The data set is retrieved from the package *City of Toronto Neighbourhood Profiles* from the year 2011. ###### We intend to use the comparison to bring awareness of inequality in income and education existing in the City of Toronto."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, include=FALSE, message=FALSE}
#Set up the environment
library(opendatatoronto)
library(dplyr)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(knitr)
library(data.table) #Useful to transponse a dataframe
library(sf)
library(tmap)
library(tmaptools)
library(leaflet)
```

# Introduction
Education is a contributor to many beneficial socio-economic outcomes. In this study, we intend to examine the relationship between household income and educational profile in terms of residents' academic majors in Toronto's 140 neighbourhoods. The majors are categorized as follows—
"visual and performing arts and communications technologies," "humanities," "social and behavioural sciences and law," "business_management_and_public_administration," "physical and life sciences and technologies," "mathematics computer and information sciences," "architecture engineering and related technologies," "agriculture natural resources and conservation," "personal protective and transportation services" and "no postsecondary certificate diploma or degree." The question we want to answer is if we can find a higher population percentage in specific academic majors in higher-income neighbourhoods or vice versa? The question is relevant because the academic major is a significant factor for people's occupations and their associated earning; thus, it is deemed as an essential tool for social mobility. 

```{r echo=FALSE, include=FALSE, message=FALSE}
#Set working Directory
#setwd("~/Experimental Design for Data Science/ProblemSet2")
# Search packages (this returns a table)
neighbourhood_packages <- search_packages("Neighbourhood")
# Filter the neighbourhood demographic dataset
neighbourhood_demographics_package<- neighbourhood_packages %>%
  filter(title == "Neighbourhood Profiles")
#Create main dataset for year 2011 
main_df_2011 <-  neighbourhood_demographics_package %>% # Start with the package 
  list_package_resources() %>% # List the resources in the package
  filter(name == "Neighbourhood Data 2001, 2006, 2011.xlsx") %>% # Only keep the resource we want 
  get_resource()
### GEO-LOCATION DATASET TORONTO BY NEIGHBOURHOODS ###
# get package
package <- show_package("4def3f65-2a65-4a4f-83c4-b2a4aed72d46")
package
# get all resources for this package
resources <- list_package_resources("4def3f65-2a65-4a4f-83c4-b2a4aed72d46")
# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
# load the first datastore resource as a sample
geo_data <- filter(datastore_resources, row_number()==1) %>% get_resource()
```

# Dataset and method descriptiion
City of Toronto Neighbourhood Profiles data set is sourced from several Census tables released by Statistics Canada every five years. The dataset uses this Census data to provide a portrait of the demographic, social and economic characteristics of the people and households in each City of Toronto neighbourhood. Each data point in this file is presented for the City's 140 neighbourhoods, as well as for the City of Toronto as a whole. The data set consists of 144 columns for neighbourhoods and 1537 rows for characteristics of the neighbourhoods. Data cleaning and reshaping are performed to have a new data frame that has 14 characteristics regarding topics of income as columns and education and the 140 neighbourhoods as rows.

Exploratory data analysis is conducted to obtain insight from the dataset. To get a picture of a neighbourhood's economic status, we look at the median total household income which divides the income distribution into two equal groups, half having income above that amount, and half having income below that amount. As for the education profile, the academic majors of the residents are the variables in which we are interested. Each field for the major is the number of people who took this major in the neighbourhood. To be able to compare the neighbourhood regardless of the difference in population, we turn the value to percentages by dividing the total population between 25 to 64 years old by the population in each major. 

## Limitation of the dataset and approach
Because we chose to look at the median instead of the mean of household income, we are excluding the outliers in the income distribution. The median value would give us a more accurate picture of the neighbourhoods' economic status but also prohibit us from investigating further the outliers that are skewing the distributions. Moreover, although there are datasets for 2001,2006, 2011 and 2016, the structure and available information of the datasets are inconsistent. For example, median total household income for all neighbourhoods is only available in the dataset from 2011; therefore, this study is limited to the year 2011 and is unable to provide a comparison between Census years regarding the changes in income and education profile of the neighbourhoods. Finally, we are also not able to find out the income distribution by academic fields. Therefore, earnings differences linking to differences in residents' fields of study are not included in our analysis. This prevents from knowing if specific academic majors are more significant contributors to the neighbourhoods' total household income.

## Ethical considerations
Population percentage in specific majors have correlations with median household income in the 140 neighbourhoods. For example, higher income neighbourhoods have more percentage of business majors, and lower income neighbourhoods have more percentage of transportation majors. This observation, without further anaylysis, potentially projects social status and ranking among academic fields. 

# Income distribution map
Darker areas in the above map indicate high levels of unemployment, while lighter areas indicate low levels of unemployment. As Figure 1 shows, we observe that higher income neighbourhoods tend to be in the central part of Toronto, neighbouring each other.

# Correlation between income and academic majors
We conducted a correlation matrix table to see if certain majors have correlation with household income. As a result, we find that business, education, and [sss] have stronger positive correlations, which no certificate has strong negative correlation. 

# Majors distribution map
The four categories that have the largest correlation scores are plotted into choropleth maps to be compared with Figure 1. The distribution of business, education and asfd have the similar pattern with 

# Comparison between polarized neighbourhoods and the City of Toronto

```{r echo=FALSE, include=FALSE, message=FALSE}
##### Distribution by education 2011####
#Select sheet 2011
clean_header_df_2011 <- main_df_2011$`2011`
#Clean header using janitor
clean_header_df_2011 <- janitor::clean_names(clean_header_df_2011)
#Select only Education by major from the main dataset 
education_df_2011 <- filter(clean_header_df_2011, category == "Education" & topic == "Major field of study - Classification of Instructional Programs (CIP)")
# Remove row with the totals 
education_df_2011 <-
  education_df_2011 %>% 
    filter(!str_detect(attribute, "Total"))
#Remove first two columns 
education_df_2011 <- education_df_2011[,3:length(education_df_2011)]
#Change class of columns containing numbers as.numeric 
education_df_2011[,2:length(education_df_2011)] <- sapply(education_df_2011[,2:length(education_df_2011)], function(x) as.numeric(x))
```
```{r echo=FALSE, include=FALSE, message=FALSE}
#Create new dataframe
scaled_education_df_2011 <- education_df_2011
#Scale dataframe by column
scaled_education_df_2011 <-
  scaled_education_df_2011 %>%
    janitor::adorn_percentages(denominator = "col")
#Multiply by 100
scaled_education_df_2011[,2:length(education_df_2011)] <- sapply(scaled_education_df_2011[,2:length(scaled_education_df_2011)], function(x) (x*100))
#Round new scales to 2 digits
scaled_education_df_2011 <-
  scaled_education_df_2011 %>%
    janitor::adorn_rounding(digits = 2)
#Transform the result to dataframe
scaled_education_df_2011 <- as.data.frame(scaled_education_df_2011)
```
```{r echo=FALSE, include=FALSE, message=FALSE}
#Save neighbourhoods into a dataframe
col_names_2011 <- as.data.frame(colnames(scaled_education_df_2011))
col_names_2011 <- col_names_2011[2:nrow(col_names_2011),]
col_names_2011 <- as.data.frame(col_names_2011)
colnames(col_names_2011) <- "neighbourhoods"
col_names_2011
#Reshape dataframe
edu_df_2011_reshaped <- transpose(scaled_education_df_2011, make.names=1)
#Include neighbourhoods column
edu_df_2011_reshaped <- mutate(edu_df_2011_reshaped, neighbourhoods = col_names_2011$neighbourhoods)
#Clean header names
edu_df_2011_reshaped <- janitor::clean_names(edu_df_2011_reshaped)
#Reorder columns
edu_df_2011_reshaped <- select(edu_df_2011_reshaped, neighbourhoods, no_postsecondary_certificate_diploma_or_degree:other_fields_of_study)
```


```{r echo=FALSE, include=FALSE, message=FALSE}
##### Distribution by income 2011####
#Select sheet 2011
clean_income_df_2011 <- main_df_2011$`2011`
#Clean header using janitor
clean_income_df_2011 <- janitor::clean_names(clean_income_df_2011)
#Select only median income from the main dataset 
income_df_2011 <- clean_income_df_2011[756,]
#Remove first two columns 
income_df_2011 <- income_df_2011[,3:length(income_df_2011)]
#Change class of columns containing numbers as.numeric 
income_df_2011[,2:length(income_df_2011)] <- sapply(income_df_2011[,2:length(income_df_2011)], function(x) as.numeric(x))
```
```{r echo=FALSE, include=FALSE, message=FALSE}
#Create new dataframe
col_names_2011_income  <- as.data.frame(colnames(income_df_2011))
#Save neighbourhoods into a dataframe
col_names_2011_income <- col_names_2011_income[2:nrow(col_names_2011_income),]
col_names_2011_income <- as.data.frame(col_names_2011_income)
colnames(col_names_2011_income) <- "neighbourhoods"
#Reshape dataframe
inco_df_2011_reshaped <- transpose(income_df_2011, make.names=1)
#Include neighbourhoods column
inco_df_2011_reshaped <- mutate(inco_df_2011_reshaped, neighbourhoods = col_names_2011_income$neighbourhoods)
#Clean header names
inco_df_2011_reshaped <- janitor::clean_names(inco_df_2011_reshaped)
#Reorder columns
inco_df_2011_reshaped <- select(inco_df_2011_reshaped, neighbourhoods, median_household_total_income)
```
```{r}
### Cleaning geo-location dataset
clean_geo_data <- janitor::clean_names(geo_data)
clean_geo_data <- extract(clean_geo_data, area_name, into = "neighbourhoods" , regex = "([^(0-9)]+)")
clean_geo_data["neighbourhoods"] <- 
  janitor::make_clean_names(as.matrix(clean_geo_data["neighbourhoods"]))
clean_geo_data <- janitor::clean_names(clean_geo_data)
clean_geo_data <- select(clean_geo_data, neighbourhoods, longitude, latitude, geometry)
filter(clean_geo_data, neighbourhoods %in% c("mimico","weston_pellam_park"))
clean_geo_data["neighbourhoods"][c(17,67),] <- c("mimico_includes_humber_bay_shores","weston_pelham_park")
clean_geo_data
```
# Income and Education by Major 2011


```{r echo=FALSE, include=FALSE, message=FALSE}
#Merge income and education 2011
merged_df <- merge(edu_df_2011_reshaped, inco_df_2011_reshaped, by = 'neighbourhoods')
merged_df <- merge(merged_df, clean_geo_data, by = 'neighbourhoods', all.x = TRUE)
#Create a sf version
sf_merged_df <- st_sf(merged_df, sf_column_name = "geometry")
colnames(sf_merged_df) <- c("neighbourhoods" ,
"no_certificate" ,
"education" ,
"arts_&_com",
"humanities" ,
"social_&_law" ,
"business" ,
"physical_&_life" ,
"mathematics_&_it" ,
"architecture_engin" ,
"agriculture" ,
"health" ,
"transportation" ,
"other" ,
"median_income",
"longitude",                                                 
"latitude",                                                                                  "geometry")
colnames(merged_df)
```
```{r echo=FALSE, include=TRUE, message=FALSE, fig.height = 1.5, fig.width = 9}
### Explore relationship between income and every major percentage ###
#Create a new dataframe and change column names. Remove last three columns
education_percentage_only <- merged_df[,1:(length(merged_df)-3)]
colnames(education_percentage_only) <- c("neighbourhoods" ,
"no_certificate" ,
"education" ,
"arts_&_com",
"humanities" ,
"social_&_law" ,
"business" ,
"physical_&_life" ,
"mathematics_&_it" ,
"architecture_engin" ,
"agriculture" ,
"health" ,
"transportation" ,
"other" ,
"median_income")
#Plot each major in terms of income (y) and neighbourhood participation in each major (x)
education_percentage_only %>%
  gather(-c(median_income, neighbourhoods), key = "var", value = "value") %>% 
  ggplot(aes(x = value, y = median_income)) +
    #facet_wrap(~ var, scales = "free") +
    facet_grid(~ var, ) +
    geom_point(shape=20, color="blue", size=1) +
    stat_smooth(method=lm, se=FALSE,  colour="red") +
    labs(title = "Major by Income", x = "Neighbourhood Distribution by Major (%)", 
       y = "Median Income")
```

```{r echo=FALSE, include=TRUE, message=FALSE, fig.height = 5, fig.width = 9}
### Business major seem to have a stiff line, lets zoom in to it ###
# Get highest 5 and lowest 5 income values
income <- education_percentage_only$median_income
h5th <- sort(income)[length(income)-4]
l5th <- sort(income)[5]
# Assign income labels to the neighbourhoods
education_percentage_only$range<-ifelse(income >= h5th,"top 5",
		ifelse(income <= l5th,"bottom 5","rest"
		))
# Plot business major by income and neighbourhood distribution
ggplot(education_percentage_only, aes(x=business, y=median_income, color=range)) +        geom_point() + 
    labs(title = "Business Major by Income", x = "Distribution by Major (%)", 
       y = "Median Income")
# Run linear regression model
linearMod <- lm(business ~ median_income, data=education_percentage_only)
summary(linearMod)
```


```{r echo=FALSE, include=TRUE, message=FALSE, fig.height = 4, fig.width = 9}
### Analizing the highest income neighbourhood ###
# Plot major percentage distribution in highest income neighbourhood
education_df_highest <- education_percentage_only %>%
  filter(median_income == max(median_income) )
data_plot_highest <- 
  education_df_highest %>% 
  pivot_longer(cols = "education":"transportation", names_to = "major")
# Make a bar chart
data_plot_highest %>% 
  ggplot(aes(x = major, y = value, fill = major)) +
  geom_col()+
  ylim(0,30)+
  scale_x_discrete(
        labels=c("Agri", "Arch", "Art", "Bus","Edu", "Health", "Hum","Math", "Phys","Social", "Tran"))+
  labs(title = "Highest Income Neighbourhood by Major", x = "Major", 
       y = "Distribution (%)")
```

```{r echo=FALSE, include=TRUE, message=FALSE, fig.height = 4, fig.width = 9}
### Analizing the lowest income neighbourhood ###
# Plot major percentage distribution in lowest income neighbourhood
education_df_highest <- education_percentage_only %>%
  filter(median_income == min(median_income) )
data_plot_highest <- 
  education_df_highest %>% 
  pivot_longer(cols = "education":"transportation", names_to = "major")
# Make a bar chart
data_plot_highest %>% 
  ggplot(aes(x = major, y = value, fill = major)) +
  geom_col()+
  ylim(0,30)+
  scale_x_discrete(
        labels=c("Agri", "Arch", "Art", "Bus","Edu", "Health", "Hum","Math", "Phys","Social", "Tran"))+
  labs(title = "Lowest Income Neighbourhood by Major", x = "Major", 
       y = "Distribution (%)")
```



```{r echo=FALSE, include=TRUE, message=FALSE, fig.align ="right", fig.height = 5, fig.width = 9, warning=FALSE}
#Plot Median Income
tmap_mode("plot")
tm_shape(sf_merged_df) +
  tm_polygons(col = "median_income",
              breaks = c(25000,50000,75000,100000,125000,150000,200000),
              #style = "quantile",
              legend.hist = TRUE,
              palette = "seq") +
  tm_layout(legend.outside = TRUE)
#Plot all majors without palette
tmap_mode("plot")
tm_shape(sf_merged_df) +
tm_scale_bar(width = 0.5)+
tm_layout(legend.show = TRUE, legend.position = c("right", "bottom"), title.size = 2, title.position = c("center","center")) +
  tm_polygons(c("no_certificate" ,
"education" ,
"arts_&_com",
"humanities" ,
"social_&_law" ,
"business" ,
"physical_&_life" ,
"mathematics_&_it" ,
"architecture_engin" ,
"agriculture" ,
"health" ,
"transportation")) +
  tm_facets(sync = TRUE, ncol = 4)
```


```{r echo=FALSE, include=TRUE, message=FALSE, warning=FALSE}
correlation_df <- as.data.frame(cor(education_percentage_only[,2:(length(education_percentage_only)-2)], education_percentage_only$median_income))
correlation_df <- round(correlation_df, digits = 2)
colnames(correlation_df) <- "correlation"
correlation_df <- mutate(correlation_df, major = rownames(correlation_df))
#Reorder columns
correlation_df <- select(correlation_df, major, correlation)
correlation_df <- correlation_df[order(correlation_df$correlation),]
row.names(correlation_df)<-1:nrow(correlation_df)
kable(correlation_df)
```